---
title: "Phonology of East Caucasian languages"
author: "G. Moroz"
output: html_document
bibliography: "bibliography.bib"
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# setwd("/home/agricolamz/work/materials/2021_caucasian_phonology")
library(lingtypology)
library(tidyverse)
df <- read_csv("database.csv")
# for maps
df %>% 
  mutate(lang = str_replace_all(lang, "Khwarshi", "Khwarshi-Inkhoqwari"),
         lang = str_replace_all(lang, "Dargwa", "Megeb")) ->
  df
```


## Introduction

There are a lot of work about phonology of indigenous languages of the Caucasus [@catford77; @smeets94a; @smeets94b; @alekseev01; @hewitt04; @grawunder17; @begus21; @boris21b; @boris21a; @koryakov21] and particular dedicated East Caucasian languages [@kk90]. There are also a lot of work on historical-comparative phonetics (S. M. Gasanova, G. Kh. Ibragimov, P. T. Magomedova, U. A. Meylanova, B. B. Talibov, S. M. Khaydakov, A. E. Alekseev, Ju. D. Desheriev, B. K. Gigineishvili, D. S. Imnaishvili, J. Nichols, S. L. Nikolayev, S. A. Starostin, A. S. Kasyan, N. A. Ardoteli, O. A. Mudrak, PUT HERE BIBTEX). Fortunately the amount of grammatical description of particular languages is increasing, so we have a lot of more new and detailed information about phonological inventories of particular villages and do not need to extropolate our knowledge of standard languages onto all villages, where the language is spoken. Even though we have a lot of works on different East Caucasian languages, in order to proceed with the analysis we still need a unified description of those inventories. We compiled the database of East Caucasian languages that can be downloaded here (add link). The database has the following variables:

* id -- unique id for each segment
* glottocode -- language identifier after Glottolog database [@hammarstrom21]
* lang -- language name
* idiom
* type
* value --- segment in IPA notation
* value_source --- segment as it was presented in the source
* annotation --- annotation of type of the segment (consonant vs. vowel)
* source --- bibtex entry of the source of the information
* page
* comment

As we discovered on average East Caucasian languages have more consonants and vowels then other languages of the world. The main reason for this are the folowing

* as other indigenous languages of the Caucasus all East Caucasian languages have **ejective consonants** (except Udi);
* as other indigenous languages of the Caucasus all East Caucasian languages have **uvular consonants**;
* Andic and Tsezic languages have **lateral obstruents**;
* **labialization**, gemination and fortis/lenis destinction is widespread across Dagestan;
* common among languages of the world triangle vowel system (*i*, *e*, *a*, *o*, *u*) is complicated with **nasalization** (mostly Andic and some Tsezic languages), **long vowels** (Nakh, Andic and Tsezic), **pharyngealisation** and **umlaut vowels** (on the South, Azerbaijanian influence?);
* more or less East Caucasian languages share the same inventory of sonorants (*r*, *l*, *m*, *n*, *j*, *w*) with rare exception for *w* in South Lezgian languages.

## Inventory size

Here is the map of inventory sizes of East Caucasian languages:

```{r}
df %>% 
  count(lang) ->
  for_map

map.feature(languages = for_map$lang,
            minichart = "pie",
            minichart.data = for_map$n,
            minichart.labels = TRUE,
            width = 3,
            tile = "Stamen.Terrain")
```

As we see inventory size are differe from `r min(range(for_map$n))` (Lezgian and Tsez) to `r max(range(for_map$n))` (Northern Akvakh). We can compare the obtained numbers with the PHOIBLE database [@phoible] which contains inventories of the languages of the world:

```{r}
ph <- phoible.feature()
set.seed(42)
ph %>% 
  count(language, inventoryid) %>% 
  group_by(language) %>% 
  sample_n(1) %>% 
  anti_join(for_map) %>% 
  ggplot(aes(n))+
  geom_histogram(data = for_map, aes(x = n, y = ..density..))+
  geom_density(size = 2, color = "red")+
  theme_minimal()+
  labs(x = "number of segments",
       subtitle = "Histogram of East Caucasian inventory sizes (in black)\nwith overlayed density function of inventory sizes of languages of the world (in red)")
```

As we can see, East Caucasian languages in general have generally big segment inventories comparing to the rest languages of the world (we removed East Caucasian languages from PHOIBLE before creating the density curve).


## Consonant inventory size

We can do the same comparison for the consonant inventories:

```{r}
df %>% 
  filter(segment_type == "consonant") %>% 
  count(lang) ->
  for_map

map.feature(languages = for_map$lang,
            minichart = "pie",
            minichart.data = for_map$n,
            minichart.labels = TRUE,
            width = 3,
            tile = "Stamen.Terrain")
```

As we see inventory size are differe from `r min(range(for_map$n))` (Budukh) to `r max(range(for_map$n))` (Northern Akvakh). We can compare the obtained numbers with the PHOIBLE database [@phoible]:

```{r}
set.seed(42)
ph %>% 
  filter(segmentclass == "consonant") %>% 
  count(language, inventoryid) %>% 
  group_by(language) %>% 
  sample_n(1) %>% 
  anti_join(for_map) %>% 
  ggplot(aes(n))+
  geom_histogram(data = for_map, aes(x = n, y = ..density..))+
  geom_density(size = 2, color = "red")+
  theme_minimal()+
  labs(x = "number of segments",
       subtitle = "Histogram of East Caucasian consonant inventory sizes (in black)\nwith overlayed density function of consonant inventory sizes of languages of the world (in red)")
```

As we can see, the majority of languages from PHOIBLE has less consonants then East Caucasian languages.

## Vowel inventory size

We can do the same comparison for the consonant inventories:

```{r}
df %>% 
  filter(segment_type == "vowel") %>% 
  count(lang) ->
  for_map

map.feature(languages = for_map$lang,
            minichart = "pie",
            minichart.data = for_map$n,
            minichart.labels = TRUE,
            width = 3,
            tile = "Stamen.Terrain")
```

As we see inventory size are differe from `r min(range(for_map$n))` (Avar) to `r max(range(for_map$n))` (Bezhta). We can compare the obtained numbers with the PHOIBLE database [@phoible]:

```{r}
set.seed(42)
ph %>% 
  filter(segmentclass == "vowel") %>% 
  count(language, inventoryid) %>% 
  group_by(language) %>% 
  sample_n(1) %>% 
  anti_join(for_map) %>% 
  ggplot(aes(n))+
  geom_histogram(data = for_map, aes(x = n, y = ..density..))+
  geom_density(size = 2, color = "red")+
  theme_minimal()+
  labs(x = "number of segments",
       subtitle = "Histogram of East Caucasian vowel inventory sizes (in black)\nwith overlayed density function of vowel inventory sizes of languages of the world (in red)")
```

As we can see, even vowel inventory sizes of East Caucasian languages are bigger than the average.

## Labialization

```{r}
df %>% 
  mutate(lab = str_detect(segment, "ʷ")) %>% 
  group_by(lang) %>% 
  mutate(page = str_c(unique(page), sep = ", ", collapse = ", ")) %>% 
  group_by(lang, language, idiom, type, source, page, contributer, date) %>% 
  summarise(n_lab = sum(lab),
            labialization = ifelse(n_lab > 0, "attested", "not attested")) %>% 
  select(lang, language, idiom, type, labialization, source, page, contributer, date) ->
  for_map

for_map %>% 
  ungroup() %>% 
  select(-lang) %>% 
  write_csv("for_dagatlas/labialization.csv", na = "")

map.feature(languages = for_map$lang,
            features = for_map$labialization,
            tile = "Stamen.Terrain")
```

Labialization is widely attested in East Caucasian languages, however there are some languages that lost this feature (e. g. in Budukh, [@talibov07]). Unfortunately, sometimes scholars do not put the whole list of labialised consonants attested in a language.

## Geminates/Fortis

```{r}
df %>% 
  filter(segment_type == "consonant") %>% 
  mutate(feature = str_detect(segment, "ː")) %>% 
  group_by(lang) %>% 
  mutate(page = str_c(unique(page), sep = ", ", collapse = ", ")) %>% 
  group_by(lang, language, idiom, type, source, page, contributer, date) %>% 
  summarise(n_feature = sum(feature),
            geminates = ifelse(n_feature > 0, "attested", "not attested")) %>% 
  select(lang, language, idiom, type, geminates, source, page, contributer, date) ->
  for_map

for_map %>% 
  ungroup() %>% 
  select(-lang) %>% 
  write_csv("for_dagatlas/geminates.csv", na = "")

map.feature(languages = for_map$lang,
            features = for_map$geminates,
            tile = "Stamen.Terrain")
```

## Lateral obstruents

```{r}
df %>% 
  mutate(feature = str_detect(segment, "ɬ")) %>% 
  group_by(lang) %>% 
  mutate(page = str_c(unique(page), sep = ", ", collapse = ", ")) %>% 
  group_by(lang, language, idiom, type, source, page, contributer, date) %>% 
  summarise(n_feature = sum(feature),
            laterals = ifelse(n_feature > 0, "attested", "not attested")) %>% 
  select(lang, language, idiom, type, laterals, source, page, contributer, date) ->
  for_map

for_map %>% 
  ungroup() %>% 
  select(-lang) %>% 
  write_csv("for_dagatlas/laterals.csv", na = "")

map.feature(languages = for_map$lang,
            features = for_map$laterals,
            tile = "Stamen.Terrain")
```


## Labial approximmant vs fricatives

```{r}
df %>% 
  mutate(feature = str_detect(segment, "v")) %>% 
  group_by(lang) %>% 
  mutate(page = str_c(unique(page), sep = ", ", collapse = ", ")) %>% 
  group_by(lang, language, idiom, type, source, page, contributer, date) %>% 
  summarise(n_feature = sum(feature),
            labials = ifelse(n_feature > 0, "v", "w")) %>% 
  select(lang, language, idiom, type, labials, source, page, contributer, date) ->
  for_map

for_map %>% 
  ungroup() %>% 
  select(-lang) %>% 
  write_csv("for_dagatlas/labials.csv", na = "")


map.feature(languages = for_map$lang,
            features = for_map$labials,
            tile = "Stamen.Terrain")
```


## Pharyngealization

```{r}
df %>%  
  mutate(feature = str_detect(segment, "ˤ")) %>% 
  group_by(lang) %>% 
  mutate(page = str_c(unique(page), sep = ", ", collapse = ", ")) %>% 
  group_by(lang, language, idiom, type, source, page, contributer, date) %>% 
  summarise(n_feature = sum(feature),
            pharyngealization = ifelse(n_feature > 0, "attested", "not attested")) %>% 
  select(lang, language, idiom, type, pharyngealization, source, page, contributer, date) ->
  for_map

for_map %>% 
  ungroup() %>% 
  select(-lang) %>% 
  write_csv("for_dagatlas/pharyngealization.csv", na = "")

map.feature(languages = for_map$lang,
            features = for_map$pharyngealization,
            tile = "Stamen.Terrain")
```

Pharyngealization --- is a complex feature that phonetically reallised as a constriction in the pharynx or epiglottis zone. Different scholars used different strategy in describing locus of the pharyngealisation: sometimes it is analysed as a vowel feature [@forker13], sometimes as a consonant feature [@khalilova09] and sometimes as a suprasegmental feature [@kibrik1994a; @moroz19]. It is common that pharyngealisation spreads through the word, hovewer most researchers distinguish core pharyngealised syllable [@kibrik1994a; @talibov07; @moroz19; @belyayev21]. Everything become even more complicated since some East Caucasian languages developed two different types of pharyngealisation [e. g. @belyayev21]. As we can see from the map, pharyngealization is common for central and south Daghestan for languages of Dargic, Lak, Tsezic and Lezgian branches.

## Nasalized vowels

```{r}
df %>% 
  filter(segment_type == "vowel") %>% 
  mutate(feature = str_detect(segment, "ⁿ")) %>% 
  group_by(lang) %>% 
  mutate(page = str_c(unique(page), sep = ", ", collapse = ", ")) %>% 
  group_by(lang, language, idiom, type, source, page, contributer, date) %>% 
  summarise(n_feature = sum(feature),
            nasalization = ifelse(n_feature > 0, "attested", "not attested")) %>% 
  select(lang, language, idiom, type, nasalization, source, page, contributer, date) ->
  for_map

for_map %>% 
  ungroup() %>% 
  select(-lang) %>% 
  write_csv("for_dagatlas/nasalization.csv", na = "")

map.feature(languages = for_map$lang,
            features = for_map$nasalization,
            tile = "Stamen.Terrain")
```

## Long vowels

```{r}
df %>% 
  filter(segment_type == "vowel") %>% 
  mutate(feature = str_detect(segment, "ː")) %>% 
  group_by(lang) %>% 
  mutate(page = str_c(unique(page), sep = ", ", collapse = ", ")) %>% 
  group_by(lang, language, idiom, type, source, page, contributer, date) %>% 
  summarise(n_feature = sum(feature),
            `long vowels` = ifelse(n_feature > 0, "attested", "not attested")) %>% 
  select(lang, language, idiom, type, `long vowels`, source, page, contributer, date) ->
  for_map

for_map %>% 
  ungroup() %>% 
  select(-lang) %>% 
  write_csv("for_dagatlas/long_vowels.csv", na = "")

map.feature(languages = for_map$lang,
            features = for_map$`long vowels`,
            tile = "Stamen.Terrain")
```


## Umlaut vowels

```{r}
df %>% 
  filter(segment_type == "vowel") %>% 
  mutate(feature = str_detect(segment, "[yøæ]")) %>% 
  group_by(lang) %>% 
  mutate(page = str_c(unique(page), sep = ", ", collapse = ", ")) %>% 
  group_by(lang, language, idiom, type, source, page, contributer, date) %>% 
  summarise(n_feature = sum(feature),
            `umlaut vowels` = ifelse(n_feature > 0, "attested", "not attested")) %>% 
  select(lang, language, idiom, type, `umlaut vowels`, source, page, contributer, date) ->
  for_map

for_map %>% 
  ungroup() %>% 
  select(-lang) %>% 
  write_csv("for_dagatlas/umlaut_vowels.csv", na = "")

map.feature(languages = for_map$lang,
            features = for_map$`umlaut vowels`,
            tile = "Stamen.Terrain")
```
